---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
```

# Academic bias

I sometimes find people claiming there is no political bias in academia, or
that it isn't a big deal. So I made these graphs to help visualize the
scale of the issue.

```{r setup}

#== Data ====
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggtext)
library(ggrepel)
library(glue)

cp <- readr::read_csv("dataverse_files/countypres_2000-2020.csv")
cpl <- cp |> 
        filter(mode == "TOTAL", year == 2020) |> 
        tidyr::pivot_wider(id_cols = -candidate, names_from = party, 
                           values_from = candidatevotes) |> 
        mutate(
          dem_share = DEMOCRAT/totalvotes,
          dem_rep   = DEMOCRAT/(DEMOCRAT+REPUBLICAN),
          dem_rep_other = dem_rep
        )

copop <- readr::read_csv("co-est2022-alldata.csv") |> 
         filter(SUMLEV == "050") |> 
         select(
           STATE, COUNTY, pop = POPESTIMATE2020
         ) |> 
         mutate(
           county_fips = paste0(STATE, COUNTY),
           pop_share = pop/sum(pop)
         )
cpl <- cpl |> 
         left_join(copop, by = "county_fips", relationship = "one-to-many") |> 
         filter(! is.na(pop))

acad <- readr::read_csv("academics-by-discipline.csv", col_types = "cnnnnnc") |> 
          rename(Discipline = `...1`) |> 
          filter(Discipline != "Total") |> 
          rowwise() |> 
          mutate(
            Discipline = gsub("/.*", "", Discipline),
            dem_rep = Democratic/(Democratic + Republican),
            prop_more_dem = sum(cpl$pop_share[cpl$dem_rep > dem_rep]),
            Other = `Not Affiliated` + `Not Registered`,
            dem_rep_other = (Democratic + Other/2)/
              (Democratic + Republican + Other),
            prop_more_dem_other = sum(cpl$pop_share[cpl$dem_rep > dem_rep_other]),
          )

dens_data <- density(cpl$dem_rep, weights = cpl$pop/sum(cpl$pop))
dens_data <- as.data.frame(dens_data[1:2])
acad$density_y <- sapply(acad$dem_rep, function (x) {
  dens_data$y[which.min(abs(x - dens_data$x))]
})
acad$density_y_other <- sapply(acad$dem_rep_other, function (x) {
  dens_data$y[which.min(abs(x - dens_data$x))]
})
acad$colour <- palette("Dark 2")[2:6]


libarts <- tribble(
  ~Discipline, ~dem_rep,
  "Engineering", 1.6,
  "Chemistry", 5.2, 
  "Economics", 5.5, 
  "Professional", 5.5,
  "Mathematics", 5.6,
  "Physics", 6.2,
  "Computers", 6.3,
  "Polisci", 8.2,
  "Psychology", 16.8,
  "History", 17.4,
  "Philosophy", 17.5,
  "Biology", 20.8,
  "Language", 21.1,
  "Environmental", 25.3,
  "Geoscience", 27, 
  "Classics", 27.3,
  "Theatre", 29.5,
  "Music", 32.8,
  "Art", 40.3,
  "Sociology", 43.8,
  "English", 48.3,
  "Religion", 70,
  "Anthropology", 1e6, # no republicans
  "Communications", 1e6
) |> 
  rowwise() |> 
  mutate(
    dem_rep = dem_rep/(dem_rep+1),
    prop_more_dem = sum(cpl$pop_share[cpl$dem_rep > dem_rep]),
    density_y = sapply(dem_rep, function (x) {
                  dens_data$y[which.min(abs(x - dens_data$x))]
                })
  )

acad2 <- tribble(
  ~Discipline, ~dem_rep,
  "Anthropology", 42.2,
  "Biology", 9.4,
  "Chemistry", 4.5,
  "Economics", 3.0,
  "English", 26.8,
  "Maths", 5.5,
  "Philosophy", 11.4,
  "Psychology", 11.5,
  "Sociology", 27.0
) |> 
  rowwise() |> 
  mutate(
    dem_rep = dem_rep/(dem_rep+1),
    prop_more_dem = sum(cpl$pop_share[cpl$dem_rep > dem_rep]),
    density_y = sapply(dem_rep, function (x) {
                  dens_data$y[which.min(abs(x - dens_data$x))]
                })
  )

```

The density plot below shows the population-weighted distribution of US counties
by Democratic vote share in 2020. The coloured lines show the proportion of
registered Democrats, out of registered Republicans and Democrats, in five
academic disciplines. I've added a zoom inset, in case you strain your eyes
trying to spot the number of Republicans among history faculty.

```{r plot1}

subtitle <- glue("<span style='color:{acad$colour}'>{tolower(acad$Discipline)}</span>") |> 
            glue_collapse(sep = ", ", last = " and ") |> 
            paste0("Disciplines: ", ... = _, ".\n", 
              "Density is population-weighted. ",
              "Numbers: % of population in counties with a higher Democrat share."
            )
ggp <- ggplot(cpl, aes(dem_rep)) + 
  geom_density(aes(weight = pop), fill = "grey80", color = NA) +
  geom_segment(data = acad, y = 0,  
              mapping = aes(xend = dem_rep, yend = density_y,
                            color = Discipline), linewidth = 1.1) +
  geom_point(data = acad, aes(y = density_y, color = Discipline), 
             size = 2.5) +
  scale_color_manual(values = setNames(acad$colour, acad$Discipline)) +
  scale_y_continuous(breaks = NULL, expand = expansion(0, 0)) +
  theme_minimal() +
  labs(
    x = "Democrat share", 
    y = "",
    title = "Share of Democrats in academic disciplines and US counties",
    subtitle = subtitle,
    caption = "Source: Langbert, Quain, and Klein (2016); 2020 county presidential returns; Census population estimates."
  ) +
  theme(
    legend.position = "none",
    panel.grid = element_blank(),
    panel.border = element_blank(),
    plot.title = element_textbox_simple(family = "Helvetica"),
    plot.subtitle = element_textbox_simple(size = 11, lineheight = 1.2)
  )

ggp_main <- ggp + 
  geom_text(data = acad, aes(y = density_y, color = Discipline, 
                             label = round(prop_more_dem*100, 1)), 
            nudge_y = 0.12) +
  coord_cartesian(ylim = c(-0.1, 4))

ggp_zoom <- ggp + 
  geom_text(data = acad, aes(y = density_y, color = Discipline, 
                             label = round(prop_more_dem*100, 2)), 
            nudge_y = 0.03) +
  coord_cartesian(xlim = c(0.89, 0.99), ylim = c(0, 0.2)) +
  labs(title = "", subtitle = "", x = "", caption = "") +
  theme(
    plot.background = element_rect(color = "grey20"),
    plot.margin = unit(c(0,3,1,3), "mm"),
    panel.spacing = unit(0, "mm"),
    plot.caption = element_blank(),
    plot.title = element_blank(),
    plot.subtitle = element_blank(),
    axis.title = element_blank()
  )
  
grob_zoom <- ggplotGrob(ggp_zoom)


ggp_main + 
  annotation_custom(grob_zoom, xmin = 0.65, xmax = 0.97, ymin = 2.3, ymax = 3.9) +
  annotate("rect", xmin = .87, xmax = 0.99, 
           ymin = -0.05, ymax = 0.5, fill = NA,
           linewidth = 0.3, color = "grey20") +
  annotate("segment", x = c(0.87, 0.99), xend = c(0.65, 0.97), y = 0.5, yend = 2.3, 
           linewidth = 0.3, linetype = 3)


```


Another set of data comes from Langbert and Stevens (2020):

```{r plot1.1}

subtitle <- paste0(
              "Density is population-weighted. ",
              "Numbers: % of population in counties with a higher Democrat share."
            )
ggpalt <- ggplot(cpl, aes(dem_rep)) + 
  geom_density(aes(weight = pop), fill = "grey80", color = NA) +
  geom_segment(data = acad2, y = 0,  
              mapping = aes(xend = dem_rep, yend = density_y,
                            color = Discipline), linewidth = 1.1) +
  geom_point(data = acad2, aes(y = density_y, color = Discipline), 
             size = 2.5) +
  scale_y_continuous(breaks = NULL, expand = expansion(0, 0)) +
  theme_minimal() +
  labs(
    x = "Democrat share", 
    y = "",
    title = "Share of Democrats in academic disciplines and US counties",
    subtitle = subtitle,
    caption = "Source: Langbert and Stevens (2020); 2020 county presidential returns; Census population estimates."
  ) +
  theme(
    legend.position = "none",
    panel.grid = element_blank(),
    panel.border = element_blank()
  )

ggpalt_main <- ggpalt +
  geom_text(data = acad2 |> filter(dem_rep < 0.9), 
            aes(y = density_y, color = Discipline, 
                label = paste(Discipline, round(prop_more_dem*100, 1))
                ),
            angle = 45, hjust = 0,
            nudge_y = 0.12, check_overlap = TRUE) +
  coord_cartesian(ylim = c(-0.1, 4))

acad2 <- acad2 |> arrange(dem_rep)
acad2$vjust <- rep_len(c(-0.5, 0.5), nrow(acad2))
acad2$vjust[acad2$Discipline == "Anthropology"] <- 0.7
acad2$angle <- 45

ggpalt_zoom <- ggpalt + 
  geom_text(data = acad2, aes(y = density_y, color = Discipline, 
                             label = paste(Discipline, round(prop_more_dem*100, 1)),
                             vjust = vjust,
                             angle = angle),
            hjust = -0.1) +
  coord_cartesian(xlim = c(0.9, 1.06), ylim = c(0, 0.22)) +
  labs(title = "", subtitle = "", x = "", caption = "") +
  theme(
    plot.background = element_rect(color = "grey20"),
    plot.margin = unit(c(0,3,1,3), "mm"),
    panel.spacing = unit(0, "mm"),
    plot.caption = element_blank(),
    plot.title = element_blank(),
    plot.subtitle = element_blank(),
    axis.title = element_blank()
  )
  

ggpalt_main + 
  annotation_custom(ggplotGrob(ggpalt_zoom),
                    xmin = 0.67, xmax = 0.97, ymin = 2.5, ymax = 3.9) +
  annotate("rect", xmin = .87, xmax = 0.99, 
           ymin = -0.05, ymax = 0.5, fill = NA,
           linewidth = 0.3, color = "grey20") +
  annotate("segment", x = c(0.87, 0.99), xend = c(0.67, 0.97), y = 0.5, yend = 2.5, 
           linewidth = 0.3, linetype = 3)


```


To get more disciplines, we can use Langbert (2018), who looks at professors
in liberal arts colleges. I haven't added the proportion of the population in
more Democratic counties, since it's essentially zero for most of this data.
But well done to engineering: a full 
`r scales::percent(libarts$prop_more_dem[libarts$Discipline == "Engineering"], 1)`
of the US population lives in counties which are more Democratic than engineering
faculty.

```{r plot2}

libarts_pal <- c(palette("Dark 2")[2:6], rep("black", 19))
names(libarts_pal) <- libarts$Discipline

ggpla <- ggplot(cpl, aes(dem_rep)) + 
  geom_density(aes(weight = pop), fill = "grey80", color = NA) +
  geom_segment(data = libarts, y = 0,  
              mapping = aes(xend = dem_rep, yend = density_y, x = dem_rep,
                            colour = Discipline), 
              linewidth = 0.5, position = position_dodge(width = 0.002)) +
  geom_point(data = libarts, aes(y = density_y, colour = Discipline)) +
  scale_y_continuous(breaks = NULL, expand = expansion(0, 0)) +
  theme_minimal() +
  labs(
    x = "Democrat share", 
    y = "",
    title = "Share of Democrats in academic disciplines and US counties",
    subtitle = "Professors in liberal arts colleges. Density is population-weighted.",
    caption = "Source: Langbert (2018); 2020 county presidential returns; 2020 census population estimates"
  ) +
  coord_cartesian(ylim = c(0, 4.1)) +
  theme(
    legend.position = "none",
    panel.grid = element_blank(),
    panel.border = element_blank(),
    plot.title = element_textbox_simple(family = "Helvetica", lineheight = 1.2),
    plot.subtitle = element_textbox_simple(size = 11, lineheight = 1.2)
  ) 


ggpla_main <- ggpla + 
  geom_text(data = libarts |> filter(dem_rep < 0.9),
            aes(y = density_y, label = Discipline, colour = Discipline),
            nudge_y = 0.06, nudge_x = 0.00, angle = 45, hjust = 0, vjust = 1,
            # size = 0.5,
            check_overlap = TRUE)

libarts$vjust <- rep_len(c(-0.5, 0.5), nrow(libarts))
libarts$vjust[libarts$Discipline=="English"] <- 0.5
libarts$vjust[libarts$Discipline=="Environmental"] <- -0.5
libarts$vjust[libarts$Discipline=="Geoscience"] <- 0
libarts$vjust[libarts$Discipline=="Sociology"] <- -0.25
libarts$vjust[libarts$Discipline=="Theatre"] <- 0
libarts$vjust[libarts$Discipline=="Language"] <- 1
ggpla_zoom <- ggpla + 
  geom_text(data = libarts,
            aes(y = density_y, label = Discipline, colour = Discipline, 
                vjust = vjust),
            nudge_y = 0.002, nudge_x = 0.00, angle = 45, hjust = 0, 
            size = 2.8) +
  scale_color_discrete() +
  coord_cartesian(xlim = c(0.94, 1.01), ylim = c(0, 0.036)) 

ggpla_window <- ggpla_zoom + 
  theme(
    plot.title = element_blank(),
    plot.subtitle = element_blank(),
    plot.caption = element_blank(),
    plot.background = element_rect(color = "grey20"),
    axis.title = element_blank()
  )
ggpla_main + 
  annotation_custom(ggplotGrob(ggpla_window), xmin = 0.2, xmax = 1, ymin = 2.9, 
                    ymax = 4) + 
  annotate("rect", xmin = 0.94, xmax = 1, ymin = 0, ymax = 0.1, fill = NA,
           color = "grey20", linewidth = 0.3) +
  annotate("segment", x = c(0.94, 1), xend = c(0.2, 1), y = 0.1, yend = 2.9,
           linewidth = 0.3, linetype = 3)

```

# FAQ

* Is this satire?

   A little bit, yes. But the data is real.

* This isn't fair, you're comparing *registered* Democrats in faculty to 
  Democrat *voters* in the US.
  
   Short answer: dude, suppose I'm out by a factor of ten.

   Long answer: in the 2016 data, 
   we can correct for this by assuming that unregistered faculty split their vote 
   equally between Democrat and Republican. On this ridiculously generous assumption,
   our data looks like this:

```{r plot3}

subtitle <- glue("<span style='color:{acad$colour}'>{tolower(acad$Discipline)}</span>") |> 
            glue_collapse(sep = ", ", last = " and ") |> 
            paste0("Disciplines: ", ... = _, ".\n", 
            "Numbers: % of US population in counties with a higher Democrat share.", 
            " Density is population-weighted.")
ggp2 <- ggplot(cpl, aes(dem_rep_other)) + 
  geom_density(aes(weight = pop), fill = "grey80", color = NA) +
  geom_segment(data = acad, y = 0,  
              mapping = aes(xend = dem_rep_other, yend = density_y_other,
                            color = Discipline), linewidth = 1.1) +
  geom_point(data = acad, aes(y = density_y_other, color = Discipline), 
             size = 2.5) +
  scale_color_manual(values = setNames(acad$colour, acad$Discipline)) +
  scale_y_continuous(breaks = NULL, expand = expansion(0, 0)) +
  theme_minimal() +
  labs(
    x = "Democrat share", 
    y = "",
    title = "Share of Democrats in academic disciplines and US counties, assuming unregistered faculty split 50-50",
    subtitle = subtitle,
    caption = "Source: Langbert, Quain, and Klein (2016); 2020 county presidential returns; Census population estimates."
  ) +
  theme(
    legend.position = "none",
    panel.grid = element_blank(),
    panel.border = element_blank(),
    plot.title = element_textbox_simple(family = "Helvetica"),
    plot.subtitle = element_textbox_simple(size = 11, lineheight = 1.2)
  )

ggp2 + 
  geom_text(data = acad, aes(y = density_y_other, color = Discipline, 
                             label = round(prop_more_dem_other * 100)), 
            nudge_y = 0.15) +
  coord_cartesian(ylim = c(-0.1, 4))

```


# Sources

Langbert, M., Quain, A.J. and Klein, D.B., 2016. Faculty voter registration in economics, history, journalism, law, and psychology. Econ Journal Watch, 13(3), pp.422-451.

Langbert, M., 2018. Homogenous: The political affiliations of elite liberal arts college faculty. Academic Questions, 31(2), pp.186-197.

Langbert, M. and Stevens, S., 2020. Partisan registration and contributions of faculty in flagship colleges. National Association of Scholars, 17.
