---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
```

# Academic bias

I sometimes still find people claiming there is no political bias in academia.
So I made these graphs to help explain the problem.

```{r setup}

#== Data ====
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggtext)
library(ggforce)
library(glue)

cp <- readr::read_csv("dataverse_files/countypres_2000-2020.csv")
cpl <- cp |> 
        filter(mode == "TOTAL", year == 2020) |> 
        tidyr::pivot_wider(id_cols = -candidate, names_from = party, 
                           values_from = candidatevotes) |> 
        mutate(
          dem_share = DEMOCRAT/totalvotes,
          dem_rep   = DEMOCRAT/(DEMOCRAT+REPUBLICAN),
          dem_rep_other = dem_rep
        )

copop <- readr::read_csv("co-est2022-alldata.csv") |> 
         filter(SUMLEV == "050") |> 
         select(
           STATE, COUNTY, pop = POPESTIMATE2020
         ) |> 
         mutate(
           county_fips = paste0(STATE, COUNTY),
           pop_share = pop/sum(pop)
         )
cpl <- cpl |> 
         left_join(copop, by = "county_fips", relationship = "one-to-many") |> 
         filter(! is.na(pop))

acad <- readr::read_csv("academics-by-discipline.csv", col_types = "cnnnnnc") |> 
          rename(Discipline = `...1`) |> 
          filter(Discipline != "Total") |> 
          rowwise() |> 
          mutate(
            Discipline = gsub("/.*", "", Discipline),
            dem_rep = Democratic/(Democratic + Republican),
            prop_more_dem = sum(cpl$pop_share[cpl$dem_rep > dem_rep]),
            Other = `Not Affiliated` + `Not Registered`,
            dem_rep_other = (Democratic + Other/2)/
              (Democratic + Republican + Other),
            prop_more_dem_other = sum(cpl$pop_share[cpl$dem_rep > dem_rep_other]),
          )

dens_data <- density(cpl$dem_rep, weights = cpl$pop/sum(cpl$pop))
dens_data <- as.data.frame(dens_data[1:2])
acad$density_y <- sapply(acad$dem_rep, function (x) {
  dens_data$y[which.min(abs(x - dens_data$x))]
})
acad$density_y_other <- sapply(acad$dem_rep_other, function (x) {
  dens_data$y[which.min(abs(x - dens_data$x))]
})
acad$colour <- palette("Dark 2")[2:6]


libarts <- tribble(
  ~Discipline, ~dem_rep,
  "Engineering", 1.6,
  "Chemistry", 5.2, 
  "Economics", 5.5, 
  "Professional", 5.5,
  "Mathematics", 5.6,
  "Physics", 6.2,
  "Computers", 6.3,
  "Polisci", 8.2,
  "Psychology", 16.8,
  "History", 17.4,
  "Philosophy", 17.5,
  "Biology", 20.8,
  "Language", 21.1,
  "Environmental", 25.3,
  "Geoscience", 27, 
  "Classics", 27.3,
  "Theatre", 29.5,
  "Music", 32.8,
  "Art", 40.3,
  "Sociology", 43.8,
  "English", 48.3,
  "Religion", 70,
  "Anthropology", 1e6, # no republicans
  "Communications", 1e6
) |> 
  rowwise() |> 
  mutate(
    dem_rep = dem_rep/(dem_rep+1),
    prop_more_dem = sum(cpl$pop_share[cpl$dem_rep > dem_rep]),
    density_y = sapply(dem_rep, function (x) {
                  dens_data$y[which.min(abs(x - dens_data$x))]
                })
  )

```

The density plot below shows the distribution of US counties by Democratic vote 
share in 2020. The coloured lines show the proportion of registered Democrats,
out of registered Republicans and Democrats, in five
academic disciplines. I've added a zoom inset, in case you strain your eyes trying
to spot the number of Republicans among history faculty.

```{r plot1}

subtitle <- glue("<span style='color:{acad$colour}'>{tolower(acad$Discipline)}</span>") |> 
            glue_collapse(sep = ", ", last = " and ") |> 
            paste0("Disciplines: ", ... = _, ".\n", 
              "Density is population-weighted. ",
              "Numbers: % of population in counties with a higher Democrat share."
            )
ggp <- ggplot(cpl, aes(dem_rep)) + 
  geom_density(aes(weight = pop), fill = "grey80", color = NA) +
  geom_segment(data = acad, y = 0,  
              mapping = aes(xend = dem_rep, yend = density_y,
                            color = Discipline), linewidth = 1.1) +
  geom_point(data = acad, aes(y = density_y, color = Discipline), 
             size = 2.5) +
  scale_color_manual(values = setNames(acad$colour, acad$Discipline)) +
  scale_y_continuous(breaks = NULL, expand = expansion(0, 0)) +
  theme_minimal() +
  labs(
    x = "Democrat share", 
    y = "",
    title = "Share of Democrats in academic disciplines and US counties",
    subtitle = subtitle,
    caption = "Source: Langbert, Quain, and Klein (2016); 2020 county presidential returns; Census population estimates."
  ) +
  theme(
    legend.position = "none",
    panel.grid = element_blank(),
    panel.border = element_blank(),
    plot.title = element_textbox_simple(family = "Helvetica"),
    plot.subtitle = element_textbox_simple(size = 11, lineheight = 1.2)
  )

ggp_main <- ggp + 
  geom_text(data = acad, aes(y = density_y, color = Discipline, 
                             label = round(prop_more_dem*100, 1)), 
            nudge_y = 0.12) +
  coord_cartesian(ylim = c(-0.1, 4))

ggp_zoom <- ggp + 
  geom_text(data = acad, aes(y = density_y, color = Discipline, 
                             label = round(prop_more_dem*100, 1)), 
            nudge_y = 0.03) +
  coord_cartesian(xlim = c(0.89, 0.99), ylim = c(0, 0.2)) +
  labs(title = "", subtitle = "", x = "", caption = "") +
  theme(
    plot.background = element_rect(color = "grey20"),
    plot.margin = unit(c(0,3,1,3), "mm"),
    panel.spacing = unit(0, "mm"),
    plot.caption = element_blank(),
    plot.title = element_blank(),
    plot.subtitle = element_blank(),
    axis.title = element_blank()
  )
  
grob_zoom <- ggplotGrob(ggp_zoom)


ggp_main + 
  annotation_custom(grob_zoom, xmin = 0.65, xmax = 0.97, ymin = 2.3, ymax = 3.9) +
  annotate("rect", xmin = .87, xmax = 0.99, 
           ymin = -0.05, ymax = 0.5, fill = NA,
           linewidth = 0.3, color = "grey20") +
  annotate("segment", x = c(0.87, 0.99), xend = c(0.65, 0.97), y = 0.5, yend = 2.3, 
           linewidth = 0.3, linetype = 3)


```


To get more disciplines, we can use Langbert (2018), who looks at professors
in liberal arts colleges. I haven't added the proportion of the population in
more Democratic counties, since it's essentially zero for most of this data.
But well done to engineering: a full 
`r scales::percent(libarts$prop_more_dem[libarts$Discipline == "Engineering"], 1)`
of the US population lives in counties which are more Democratic than engineering
faculty.

```{r plot2}

libarts_pal <- c(palette("Dark 2")[2:6], rep("black", 19))
names(libarts_pal) <- libarts$Discipline

ggpla <- ggplot(cpl, aes(dem_rep)) + 
  geom_density(aes(weight = pop), fill = "grey80", color = NA) +
  geom_segment(data = libarts, y = 0,  
              mapping = aes(xend = dem_rep, yend = density_y, x = dem_rep,
                            colour = Discipline), 
              linewidth = 0.5, position = position_dodge(width = 0.002)) +
  geom_point(data = libarts, aes(y = density_y, colour = Discipline)) +
  scale_y_continuous(breaks = NULL, expand = expansion(0, 0)) +
  theme_minimal() +
  labs(
    x = "Democrat share", 
    y = "",
    title = "Share of Democrats in academic disciplines and US counties",
    subtitle = "Professors in liberal arts colleges. Density is population-weighted.",
    caption = "Source: Langbert (2018); 2020 county presidential returns; 2020 census population estimates"
  ) +
  coord_cartesian(ylim = c(0, 4.1)) +
  theme(
    legend.position = "none",
    panel.grid = element_blank(),
    panel.border = element_blank(),
    plot.title = element_textbox_simple(family = "Helvetica"),
    plot.subtitle = element_textbox_simple(size = 11, lineheight = 1.2)
  ) 


ggpla_main <- ggpla + 
  geom_text(data = libarts |> filter(dem_rep < 0.9),
            aes(y = density_y, label = Discipline, colour = Discipline),
            nudge_y = 0.06, nudge_x = 0.00, angle = 45, hjust = 0, vjust = 1,
            # size = 0.5,
            check_overlap = TRUE)

ggpla_zoom <- ggpla + 
  geom_text(data = libarts,
            aes(y = density_y, label = Discipline, colour = Discipline),
            nudge_y = 0.002, nudge_x = 0.00, angle = 45, hjust = 0, vjust = 1,
            size = 3.5,
            check_overlap = TRUE) +
  scale_color_discrete() +
  coord_cartesian(xlim = c(0.94, 1.03), ylim = c(0, 0.028)) 

ggpla_window <- ggpla_zoom + 
  theme(
    plot.title = element_blank(),
    plot.subtitle = element_blank(),
    plot.caption = element_blank(),
    plot.background = element_rect(color = "grey20"),
    axis.title = element_blank()
  )
ggpla_main + 
  annotation_custom(ggplotGrob(ggpla_window), xmin = 0.6, xmax = 1, ymin = 2.7, 
                    ymax = 4) + 
  annotate("rect", xmin = 0.94, xmax = 1, ymin = 0, ymax = 0.1, fill = NA,
           color = "grey20", linewidth = 0.3) +
  annotate("segment", x = c(0.94, 1), xend = c(0.6, 1), y = 0.1, yend = 2.7,
           linewidth = 0.3, linetype = 3)

```

# FAQ

* Is this satire?

   A little bit, yes. But the data is real.

* This isn't fair, you're comparing registered Democrats in faculty to 
  Democrat voters in the US.
  
   Short answer: dude, suppose I'm out by a factor of ten.

   Long answer: in the 2016 data, 
   we can correct for this by assuming that unregistered faculty split their vote 
   equally between Democrat and Republican. On this ridiculously generous assumption,
   our data looks like this:

```{r plot3}

subtitle <- glue("<span style='color:{acad$colour}'>{tolower(acad$Discipline)}</span>") |> 
            glue_collapse(sep = ", ", last = " and ") |> 
            paste0("Disciplines: ", ... = _, ".\n", 
            "Numbers: % of US population in counties with a higher Democrat share.", 
            " Density is population-weighted.")
ggp2 <- ggplot(cpl, aes(dem_rep_other)) + 
  geom_density(aes(weight = pop), fill = "grey80", color = NA) +
  geom_segment(data = acad, y = 0,  
              mapping = aes(xend = dem_rep_other, yend = density_y_other,
                            color = Discipline), linewidth = 1.1) +
  geom_point(data = acad, aes(y = density_y_other, color = Discipline), 
             size = 2.5) +
  scale_color_manual(values = setNames(acad$colour, acad$Discipline)) +
  scale_y_continuous(breaks = NULL, expand = expansion(0, 0)) +
  theme_minimal() +
  labs(
    x = "Democrat share", 
    y = "",
    title = "Share of Democrats in academic disciplines and US counties, assuming unregistered faculty split 50-50",
    subtitle = subtitle,
    caption = "Source: Langbert, Quain, and Klein (2016); 2020 county presidential returns; Census population estimates."
  ) +
  theme(
    legend.position = "none",
    panel.grid = element_blank(),
    panel.border = element_blank(),
    plot.title = element_textbox_simple(family = "Helvetica"),
    plot.subtitle = element_textbox_simple(size = 11, lineheight = 1.2)
  )

ggp2 + 
  geom_text(data = acad, aes(y = density_y_other, color = Discipline, 
                             label = round(prop_more_dem_other * 100)), 
            nudge_y = 0.15) +
  coord_cartesian(ylim = c(-0.1, 4))

```
